# Geoprocessing in R

For this part, we will use mostly {sf} capabilities. Most spatial functions (if not all) start with the `st_` (*for spatial type*) prefix like in [PostGIS](https://postgis.net/docs/manual-2.5/reference.html).

We will focus on vector data from the {spData} package.

```{r libraries}
library(dplyr)
library(sf)
library(spData)
library(here)
```


## Read / write
{sf} provides the `st_read` and `st_write` functions to access geospatial files. They can operate with any [vector driver provided by GDAL](https://gdal.org/drivers/vector/index.html).

We will use the data from the {spData}^[More details on the datasets here : [https://cran.r-project.org/web/packages/spData/spData.pdf](https://cran.r-project.org/web/packages/spData/spData.pdf)] package. Let's see what it contains :

```{r spData_files}
list.files(system.file("shapes", package = "spData"))
```

We will work on cycle hires in London so let's get that data.

### Cycle hire dataset

```{r load_hire_data}
cycle_hire <- st_read(system.file("shapes/cycle_hire.geojson", package="spData"))
```

Here we can see a couple of functions :

* `st_read()` is the reading function from {sf}
* `system.file()` is a function that allow us to look for data in packages, independently of the operating system.

By default, {sf} provides informations when loading the dataset. We can see it contains 742 features and 5 fields. It is points with lat/lon coordinates (we can see it through dataset SRID : 4326). {sf} also provides the bounding box and if possible the proj4string.

Here is the description of the dataset we can find in the documentation: 

```{block2, type='rmdnote'}
**cycle_hire dataset**
  
~Description:~ Points representing cycle hire points accross London.

~Format:~
  
 *  **id** Id of the hire point
 *  **name** Name of the point
 *  **area** Area they are in
 *  **nbikes** The number of bikes currently parked there  
 *  **nempty** The number of empty places
 *  *geometry* sfc_POINT

~Source~: [cyclehireapp.com/cyclehirelive/cyclehire.csv](cyclehireapp.com/cyclehirelive/cyclehire.csv)
```


We can see that there is the bike parked, the count of empty slots but not the total amount of bike slots. Let's create a new `slots` column for this.

```{r create_bike_slots}
cycle_hire <- cycle_hire %>% 
  mutate(slots = nbikes +  nempty)
```

Let's load polygon data, in this case London's boroughs stored in the *lnd* dataset.

### Boroughs of London

This dataset is contained in a R data format so the loading is different.

```{r loading_lnd}
data(lnd) # load the dataset in memory
lnd       # call the dataset to visualise the 10 first features
```

We can see this dataset contains 33 features and 7 fields, in lat/lon coordinates too.

```{block2, type='rmdnote'}
**ldn dataset**

The boroughs of London

Description : Polygons representing large administrative zones in London

Format:
  
  *  **NAME** Borough name
  *  **GSS_CODE** Official code
  *  **HECTARES** How many hectares
  *  **NONLD_AREA** Area outside London
  *  **ONS_INNER** Office for national statistics code
  *  **SUB_2009** Empty column
  *  **SUB_2006** Empty column
  *  *geometry* sfc_MULTIPOLYGON

Source : [https://github.com/Robinlovelace/Creating-maps-in-R](https://github.com/Robinlovelace/Creating-maps-in-R)
```

In order to ease spatial calculations, let's reproject them.

## Reprojection {#reprojection}

The [Ordnance Survey National Grid](https://en.wikipedia.org/wiki/Ordnance_Survey_National_Grid) is the official one for Great Britain. Its SRID is **EPSG:27700**.

```{r data_reprojection}
cycle_hire_27700 <- cycle_hire %>%
  st_transform(crs = st_crs(27700))

london_27700 <- lnd %>%
  st_transform(crs = st_crs(27700))
```

TO reproject, we used 2 functions:
- `st_transform()` for the reprojection
- `st_crs()` to get CRS definition from EPSG code

We also use the pipe operator : `%>%`, it is useful to pipe data to another function. This is provided by the [{magrittr} package](https://cran.r-project.org/web/packages/magrittr/index.html) through {dplyr} and {sf}.

Now, we can create a quick map to visualize our data. We can use the `plot()` function to do this. This function is part of base R.

```{r quick_plot}
plot(london_27700$geometry) # we just want to plot the geometry column
plot(cycle_hire_27700$geometry, 
 col = "red",  # color
 cex = 0.5,    # size of symbol
 add = TRUE)   # important parameter to create multilayer plots
```


## Joins
We can use two ways to link those datasets together, by attributes, they share their area name (`area` and `NAME`) or spatially. For the sake of the exercice, let's do both.

### Join by attributes

Let's join them  with a inner join to see how many have corresponding

```{r inner_attribute_join}
cycle_hire_27700 %>% inner_join(
  london_27700 %>%
    st_drop_geometry(), # we don't need the geometry here
  by = c( "area" = "NAME")
  
)
```


We can see that only 33 features matched. That's poor, let's try this spatially.

### Spatial join

For this, we will try to provide a `GSS_CODE` for all  cycle hire points. We will regroup the data afterwards.

For this, we will select only the `GSS_CODE` column from `london_27700` with the `select` function from {dplyr}, the geometry will follow.

```{r spatial_join}
cycle_hire_27700 <- cycle_hire_27700 %>% st_join(london_27700 %>% select(GSS_CODE))
```

Now if we look at our dataset, there is a `GSS_CODE` column.

```{r names_cycle_hire}
names(cycle_hire_27700)
```

How many points doesn't have a GSS_code ?

```{r controle_nas}
cycle_hire_27700 %>% filter(is.na(GSS_CODE))
```

Only one, that's more better than before ! I don't know well enough London to fix this. But that is not blocking.

Now to paraphrase Anita Graser   : [[@graserAggregate]](https://anitagraser.com/2017/06/08/aggregate-all-the-things-qgis-expression-edition/)

## Aggregation
### Count

```{r aggregation_count}
cycle_hire_by_area <- cycle_hire_27700 %>%
  filter(!is.na(GSS_CODE)) %>% # remove NAs
  st_drop_geometry() %>% # let's put geometry aside
  group_by(GSS_CODE) %>%  # group data by GSS_CODE
  tally(name = "count", sort= TRUE) # Aggregate
cycle_hire_by_area
```


### Sum

```{r aggregation_sum}
cycle_hire_by_area_sum <- cycle_hire_27700 %>%
  filter(!is.na(GSS_CODE)) %>% # remove NAs
  st_drop_geometry() %>% # let's put geometry aside
  group_by(GSS_CODE) %>%  # group data by GSS_CODE
  summarise(sum = sum(nbikes), count = n()) # Aggregate
cycle_hire_by_area_sum
```

We could have use the base function `aggregate()` which works with {sf} objects.

```{r aggregate_alternative}
aggregate(cycle_hire_27700["nbikes"], by = list(cycle_hire_27700$"GSS_CODE"),
                       FUN = sum, na.rm = TRUE)
```

If we want to represents our data with proportionnal symbols, we might want to have centroids. {sf} provides two functions in order to do that:

* `st_centroid()`
* `st_point_on_surface()`

`st_point_on_surface()` provides a point randomly **in** the entry shape. That can be useful for irregular shapes where the centroid might be outside the shape.

## Centroids

```{r centroids}
boroughs_centroids <- london_27700 %>%
  select(NAME, GSS_CODE) %>% # only keep useful columns
  st_centroid()

```
You can also do buffers and other geometrical operations like [`st_union()`](https://r-spatial.github.io/sf/reference/geos_combine.html) to merge geometries

![Spatial equivalents of logical operators [@lovelace_geocomputation_2019]](images/venn-clip-1.png)


## Geometric binary predicates

{sf} provides numerous geometric binary predicates that can be used with the intersection function.

* st_intersects()
* st_disjoint()
* st_touches()
* st_crosses()
* st_within()
* st_contains()
* st_contains_properly()
* st_overlaps()
* st_equals()
* st_covers()
* st_covered_by()
* st_equals_exact()
* st_is_within_distance()

You can use it alone or with  ̀st_join()`. 

For example, if we want to the cycle hires contained in the borough of Wandsworth, we will do like this.

```{r find_E09000032}
london_27700 %>% 
  filter(NAME == "Wandsworth") %>% 
  st_contains(cycle_hire_27700)
```

That will return a list of cycle hire points id.

In contrary, if we want to find in which borough the hire point with id 614 is we need to do this :

```{r borough_of_614}
cycle_hire_27700 %>% filter(id == "614") %>% 
  st_within(london_27700) # borough at index 22
```

To get the borough data, there is some more work to do.

```{r get_borough22}
london_27700[unlist(cycle_hire_27700 %>% filter(id == "614") %>% st_within(london_27700)),]
```

## Saving results


In the first part, we saw that we can read data but we can also write it !

### Writing data

To write data, we will use the `st_write()` function.
It takes the data source name (*dsn*) as mandatory argument, {sf} will try to find the good driver from the extension (here it is *gpkg* for GeoPackage).

```{block2, type='rmdwarning'}
*st_write()* can't save non geospatial data. So we need to  join the data from cycle_hire_by_area_sum to
the boroughs first.
```

As we want to save it to  GeoPackage^[Because [GeoPackage are cool !](http://switchfromshapefile.org/#geopackage)], we also need to provide a layer name : *london_boroughs_27700*. Repeat for all data you want to save.

```{r write_geodata}

london_27700 %>% left_join(cycle_hire_by_area_sum) %>%
  st_write(
    dsn = here("foss4g_R_workshop.gpkg"),
    layer = "london_boroughs_27700", 
    layer_options = "OVERWRITE=true")

boroughs_centroids %>%
  left_join(cycle_hire_by_area_sum) %>%
  st_write(
    dsn = here("foss4g_R_workshop.gpkg"),
    layer = "boroughs_centroids_27700", 
    layer_options = "OVERWRITE=true")


cycle_hire_27700 %>%
  left_join(cycle_hire_by_area_sum) %>%
  st_write(
    dsn = here("foss4g_R_workshop.gpkg"),
    layer = "cycle_hire_27700",
    layer_options = "OVERWRITE=true")
```

```{block2, type='rmdnote'}
We used the `here()` function as it preserve the project file hierarchy. It works better in Rstudio but it is still useful with Jupyter notebooks.

The data set where joined by their GSS_CODE. You can specify the "by" statement, but for the sake of readability, it is not show here.

The `layer_options = "OVERWRITE=true"` ensure you can write on existing layer, it is optionnal.
```


```{r here}
print(here())  # print the project directory
list.files(here()) # list the files in the project directory
```


### Check data

{sf} provides a `st_layers()` function that is useful to see the content of a dataset.
 
```{r check_gpkg}
st_layers(dsn = here("foss4g_R_workshop.gpkg"))
```

Now that we have data, let's do some maps on it in the next chapter !




