# Maps are plots !
We need to visualise our data. Visualizing geospatial data often means maps and maps are plots ! (In R)
You can use plotting tools like `plot` or `ggplot2`, they all accept spatial dataframes. There is also packages to help mapping like {tmap} and {cartography}.

So let's see how we do plots in R ! First we need data.

## Prepare the work
### Libraries

```{r loading_libraries}
library(sf)
library(here)
library(dplyr)
library(ggplot2)
library(tmap)
```

### Data

Let's see what data we stored previously.

```{r read_database}
st_layers(here("foss4g_R_workshop.gpkg"))
```

We can see our 3 datasets. We can loaded them into memory.

```{r load_data}
cycle_hire_27700 <- st_read(
  dsn = here("foss4g_R_workshop.gpkg"), 
  layer = "cycle_hire_27700")

london_boroughs_27700 <- st_read(
  dsn = here("foss4g_R_workshop.gpkg"), 
  layer = "london_boroughs_27700")

boroughs_centroids_27700<- st_read(
  dsn = here("foss4g_R_workshop.gpkg"), 
  layer = "boroughs_centroids_27700")
```



## The `plot()` way

You can plot any spatial dataframe. It simply works but the syntax is no easy. We did one example in \@ref(reprojection). There is several examples online if you want to go further.

## Using {ggplot}

Since we are using a lot of {tidyverse} compatible package, it is logical to use {ggplot2} which is part of this toolset. The basic philosophy of ggplot is layers, first you create an empty plot with the data then you add a geometry (`geom_*()`) and some aesthetics (`aes()`) for that geometry. You can add multiple layers and add more informations (labels, titles). All those elements are linked by a `+` symbol. It is built on [A Layered Grammar of Graphics [@WickamGrammar]](https://www.tandfonline.com/doi/abs/10.1198/jcgs.2009.07098).

```{r ggplot}
london_boroughs_27700 %>% # pipe data to
ggplot() +                # a ggplot function
  geom_sf(                # precise that it will be a spatial geometry
    aes(                  # provide some aesthetics
      geometry = geom,    # the geometry column (usually auto detected)
      fill = count)       # we want the polygon color to change following the count
    ) -> g # store it in g

g # display g
```

We can customize our map by adding thing to it. For example a viridis color scale ([friendly to common forms of colorblindness](https://bids.github.io/colormap/)).

```{r scale_vidiris}
g <- g +
  scale_fill_viridis_c(
    guide = guide_legend(title = "Hires") # legend title
  )
g
```

We can set a theme and add titles

```{r theme}
g <- g + 
  theme_bw() + 
  ggtitle("Cycle hire points", subtitle = "in London's boroughs")
g
```


```{r ggplot_multiple_layer}
ggplot() +  geom_sf(data = london_boroughs_27700) + # add boroughs shape to the map  
  geom_sf(data = boroughs_centroids_27700, # add the boroughs centroids>
          aes(size = boroughs_centroids_27700$count), # fix size of points (by area)
          color = 'red', alpha = 1/5)+ # set points colour and transparency
           ggtitle("Cycle hire points", subtitle = "in London's boroughs") + # set the map title
             theme(legend.position = 'left') + # Legend position
             scale_size_area(name = 'Hires',max_size=10) # 0 value means 0 area + legend title
```

Well, the legend is not convincing me.Although it should work like in the [OSGeoLive R Quickstart](http://live.osgeo.org/en/quickstart/R_quickstart.html).

And if we want to had decorations (scale bar, north arrow), we need to use a another package: [{ggspatial}](https://cran.r-project.org/web/packages/ggspatial/vignettes/ggspatial.html).

Now, let's use proper packages for cartography. First {tmap} who was created before {ggplot} was able to do maps with {sf} objects. Then {cartography} who is a based on `plot()` from base R.

## Using {tmap}

{tmap} was created to help people mapping when {ggplot2} was not able to plot {sf} objects^[That arrived in 2018 with the 3.0 release : https://www.tidyverse.org/articles/2018/07/ggplot2-3-0-0/]

```{r tmap_polygons}
library(tmap)

tm_shape(london_boroughs_27700) + 
    tm_polygons("count")
```

We can see that there is a color scale for the quantity of hire points.

One cool thing with {tmap}, it is that you can use 2 modes: "plot" and "view". The "view" mode is an interactive one ! Try to go over or click on polygons !


```{r tmap_interactive}
tmap_mode("view")

tm_shape(london_boroughs_27700) + 
    tm_polygons("count")
```


You can do more complex thing of course, like side by side maps or calculations.

```{r tmap_sidemaps}
tmap_mode("plot")

tm1 <- tm_shape(london_boroughs_27700) + tm_polygons("count", convert2density = TRUE)
tm2 <- tm_shape(london_boroughs_27700) + tm_bubbles(size = "count")

tmap_arrange(tm1, tm2)
```

Let's represent 2 layers in the same time :

```{r multiple_layer}
tmap_mode("view")
tm_basemap("Stamen.Watercolor") +
tm_shape(london_boroughs_27700) + tm_polygons("count", convert2density = TRUE) + tm_bubbles(size = "count", col = "red") +
tm_tiles("Stamen.TonerLabels")
```

Now we can add decorations.

```{r tmap_decorations}
tmap_mode("plot")

tm_shape(london_boroughs_27700) + 
  tm_polygons("count", convert2density = TRUE) +
  tm_bubbles(size = "count", col = "red") +
  tm_scale_bar(position=c("left", "bottom")) +
  tm_compass(size = 2, position=c("right", "top"))
```

You can place elements more precisely by giving them position coordinates.

## With {cartography}

If you want to do maps in R, please take a look at [{cartography}](http://riatelab.github.io/cartography/docs/). It is thought for doing maps and is made by geographers. The package [vignette](https://cran.r-project.org/package=cartography/vignettes/cartography.html) is very well made and provides 11 examples to cover a lot of use cases.

This is one example:

![[Example from {cartography} vignette](https://cran.r-project.org/web/packages/cartography/vignettes/cartography.html)](images/cartography_example.png)

```{r with_cartography}
library(cartography)

# Plot the boroughs
plot(st_geometry(london_boroughs_27700), col="darkseagreen3", border="darkseagreen4",  
     bg = "lightblue1", lwd = 0.5)
# Plot symbols with choropleth coloration
propSymbolsChoroLayer(
  x = london_boroughs_27700, 
  var = "sum", 
  inches = 0.1,
  border = "grey50",
  lwd = 1,
  legend.var.pos = "topright", 
  legend.var.title.txt = "Total bike hires",
  var2 = "count",
  method = "equal", 
  nclass = 4, 
  col = carto.pal(pal1 = "sand.pal", n1 = 4),
  legend.var2.values.rnd = -2,
  legend.var2.pos = "left", 
  legend.var2.title.txt = "Number of hire points"
) 
# layout
layoutLayer(title="Bike hires in London boroughs", 
            author = "Nicolas Roelandt, FOSS4G 2019", 
            sources = "Sources: spData R package, 2019", 
            scale = 5, tabtitle = TRUE, frame = FALSE)
# north arrow
north(pos = "topleft")
```


This is a more complex object but we can detail it.
First we create a plot of the background, the boroughs of London in green with a light blue background.

Then we add a `propSymbolsChoroLayer()` who is a choroplet layer with proportionnal symbols (you can have more simple items, see the documentation). SO we need to specify 2 variables on for the symbol size (here we choose `sum`) and one for the color of the symbols. For each variable we provides some complementary parameters (color palette, classification method, title of the legend, etc.).

{cartography} also provides a layout function for the map where you can insert more information : author name, date, sources, map title, position and type of the scale bar, etc.).

Last but not least you can add a north arrow with the `north()` function.

```{block2, type='rmdwarning'}
The size of the proportionnal circle in the legend are not at the right size. This problem appears with pdf output but for png output it disappears.
```


```{block2, type='rmdnotes'}
The easiest way to use {cartography} is to find what you want to do in the vignette, copy paste the code then adapt to your data and personalize !
```

