--- 
title: "R for Geospatial Processing"
author: "Nicolas Roelandt"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "This is the training materials for the **R for Geospatial Processing** workshop at FOSS4G 2019, Bucarest (Romania)."
edit: https://github.com/Bakaniko/FOSS4G2019_Geoprocessing_with_R_workshop/edit/master/%s
---

# Prerequisites

This workshop is designed for the attendance of FOSS4G 2019. So basics knowledge in GIS is expected (simple features, projections and CRS, geometrical operations, etc.).

No knowledge of R is required. A minimal knowledge of [(R)mardown](https://rmarkdown.rstudio.com/) will be a plus to take notes.

Please install R on your system and the following libraries.


```{r needed_libraries, eval=FALSE}
install.packages("dplyr")
install.packages("sf")
install.packages("spData")
install.packages("here")
install.packages("ggplot2")
install.packages("tmap")
install.packages("cartography")
```

## About this document

This document had been written with the [R Bookdown package](https://bookdown.org/yihui/bookdown) and [Rstudio](https://www.rstudio.com). You can use the editor of your choice to follow the instruction, be careful on the path given by the {here} package.

To compile this example to PDF, you need XeLaTeX. You are recommended to install TinyTeX (which includes XeLaTeX): <https://yihui.name/tinytex/>.

You can contribute to this document on GitHub :

[https://github.com/Bakaniko/FOSS4G2019_Geoprocessing_with_R_workshop](https://github.com/Bakaniko/FOSS4G2019_Geoprocessing_with_R_workshop)

```{r setup, eval=T,echo=F,message=F,warning=F}
knitr::opts_chunk$set(cache=T,echo=T,comment="",message=F,warning = F)
# Load packages
library(knitr)
library(ggplot2)
library(here)
library(spData)
library(tidyverse)
library(sf)
library(tmap)
# library(viridis)
# library(scales)
# library(cartogram)
library(ggspatial)
# library(lwgeom)
# library(mapview)
theme_set(theme_gray())
tmap_mode("plot")
```

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
